---
title: "Data Wrangling with R"
author: "Jonathan Gilligan"
date: "August 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "data"
script_dir = "scripts"

library(pacman)
p_load(zoo, xml2, tidyverse, stringr, lubridate)

theme_set(theme_bw(base_size = 15))

# source('scripts/utils.R', chdir = T)

if (!dir.exists('data')) dir.create('data')

nashville_weather <- readRDS('data/nashville_weather.Rds')
chicago_weather <- readRDS('data/chicago_weather.Rds')
```
# Data in R

## Kinds of variables

R is capable of analyzing many different kinds of data. Some of the most
important kinds of data that we may work with are:

* **Integer data**, which represents discrete quantities, such as counting
  events or objects.

* **Real number data**, which represents quantities that can have fractional values.
  Most of the data we will work with in this course, such as temperatures, 
  altitudes, amounts of rainfall, and so forth, will be real-number data.
  This kind of data is also referred to frequently as "floating point" data
  or (for obscure reasons having to do with computer hardware) as "double"
  or "double-precision" data.

* **Character data**, which represents text. Examples include names of months,
  or categories (such as the name of a city or country). This kind of data is
  also referred to as "string data".

* **factor data**, which represents variables that can only take on
  certain discrete values. R treats factor data as a kind of augmented character
  data. 
  
    The difference between character data and factor data is that factor
    data has an explicit set of allowed values and has an integer number associated
    with each of those values. 
    
    For instance, if I have a factor variable with the allowed values "up" and 
    "down", then I could not assign it a value "left" or "right", whereas a
    character variable can be assigned any arbitrary text, such as 
    "second star to the right and straight on til morning."
    
    There are two kinds of factors: ordered and unordered. The difference is that
    the legal values for ordered factors have a specific order, so you can say 
    that one comes before or after another (or is greater than or less than another),
    whereas unordered factors don't have any natural ordering.
    
    Examples of ordered variables might be the months of the year, or the days 
    of the week, or a grouping like "small", "medium", "large", or 
    "bad", "fair", "good".
    
    Examples of unordered variables might be lists of states, gender, religion, 
    cities, sports teams, or other descriptive characteristics that don't 
    have a natural order to them.

* **date and time data**, which represents calendar dates, times of day, or 
  a combination, such as 9:37 PM on January 17, 1984.

For the most part, R handles different data types sensibly so you don't need to
worry about them, but sometimes when R is reading data in from files, or when
you want to convert one kind of variable to another, you will need to think about
these.

The most common cases where you will need to think about this is when you are 
reading data in from files. Sometimes it is ambiguous whether to treat something
from a file as character data, numerical data, or a date. In such cases, you may
need to give R guidance about how to interpret data. The functions for reading
data in from files, such as `read_csv` and `read_table` allow you to specify 
whether a given column of data in a table is integer, double-precision 
(floating point), character, date, etc.

R also provides functions for converting data. The `as.character` function
takes data that might be character, factor, or numeric, and represent it as
text (characters).

`as.numeric` or `as.integer` will allow you to convert a character variable to 
a number. For instance, `as.numeric("3.14")` converts a text variable "3.14"
into a numeric variable 3.14.

`as.integer` is very useful when we want to convert an ordered factor to an 
integer that corresponds to the order of that value. For instance, if I have
an ordered factor `f` with legal values corresponding to the months of the year 
("Jan", "Feb", ..., "Dec"), if `f` has the value `Mar`, then
`as.integer(f)` will have the integer value 3.

## Vectors, Lists, Data Frames, and Tibbles.

In statistics, you generally don't just work with one number at a time, but with
collections of numbers. R provides many ways to work with collections of numbers.

### Vectors

The simplest is a **vector**. A vector is a collection of values that are all
of the same kind: a collection of integers, a collection of floating point values,
a collection of character values, a collection of factor values, etc.

You specify vectors like this: `x = c(1, 2, 5, 9, 3, 4, 2, 7, 5)`.
You can access elements of vectors by indexing their position within the
vector, so `x[3]` will be 5 and `x[4]` will be 9.

You can also give the elements of a vector names:
`ages = c(Sam = 27, Ben = 20, Sarah = 25, Deborah = 31)` allows you to use 
`ages["Ben"]`, which will be 20.

All of the elements of a vector have to be the same kind, so
`x = c(1, 2, "three")` will not allow the vector to mix numbers and characters
and R will transform all of the values to character. The result is 
`"1", "2", "three"`, and `x[1] + x[2]` will give an error because
R doesn't know how to add two character variables. However, 
`as.numeric(x[1]) + as.numeric(x[2])` will yield 3.

### Lists

Lists are a lot like vectors, but they can contain different kinds of variables.
They can even contain lists and vectors.
`x = list(1, 2, "three", list(4, 5, 6))` has four elements. The first two are
the numbers 1 and 2; the third is the character string "three", and the fourth
is the list `(4,5,6)`.

Just as we can have named vectors, we can have named lists:
`ages = list(Sam = 27, Ben = 20, Sarah = 25, Deborah = 31)`.
There is a nice shortcut to getting the elements of a named list, using a 
dollar sign: `ages$Sam` is 27.

### Data frames and tibbles

We will not use lists very much in this class. We will use vectors a little bit,
but what we will use _a lot_ are tables of data. You have probably worked a
lot with spreadsheets and other data analysis tools that organize data in a
table with rows and columns. This is a very natural and common way to work.

R provides a structure called a `data.frame` for working with tabular data, but
the package `tidyverse` introduces an improved version of the `data.frame`
called a `tibble` (think of it as a kind of data table).

`data.frames` and tibbles have rows and columns. A row represents a set of 
quantities, such as measurements or observations, that go together in some way.
Different rows in a tibble represent different sets of these quantities.

For instance, if I am measuring the height and weight of a number of people, then 
I would have a row for each person and each row would have a column for the 
person's name or identity code, a column for their height, and a column for 
their weight. 

If I am measuring the average temperature and average precipitation
for a number of cities, then I would have a column for the city, a column for the
temperature, and a column for the precipitation.

Each column of a `data.frame` or tibble should correspond to a specific kind of 
data (integer, floating point, character, factor, date, etc.). A column is 
a kind of vector, so it has to obey the restrictions that apply to vectors.

To get some experience with tibbles, let's load a couple of data sets that
I have prepared. If you have cloned the directory for this document
(from <https://github.com/gilligan-ees-3310/lab_02_documentation>),
you can load the datasets, which contain daily weather summaries for 
Nashville and Chicago,
by running the code below:

```{r load_datasets, include=TRUE}
nashville_weather = loadRDS('data/nashville_weather.Rds')
chicago_weather = loadRDS('data/chicago_weather.Rds')
```

Here is an example of the first few rows of a tibble with weather data for 
Nashville from 
`r lubridate::year(min(nashville_weather$date))`--`r lubridate::year(max(nashville_weather$date))`:

```{r head_nashville_weather, include=TRUE}
head(nashville_weather)
```

There are 6 columns: the weather station ID for the Nashville Airport,
the date of the measurement, the daily precipitation (in millimeters),
the daily minimum and maximum temperatures (Celsius), and the name of the
location.

The tibble also shows the kind of variable that each column represents:
`id` and `location` are character data, `date` is Date data, and 
`prcp`, `tmin`, and `tmax` are double-precision floating point data 
(i.e., real numbers).

In some ways, a tibble or `data.frame` is like a names list of vectors, where each
vector is a column and its name is the name of the column. We can access individual 
columns using the dollar sign, just as with regular named lists:

```{r tibble_column, include=TRUE}
precipitation = nashville_weather$prcp

head(precipitation)
```

RStudio has a nice feature that lets you examine a tibble or `data.frame`
as though it were a spreadsheet. To examine one of R's built-in data sets,
which has data on hurricanes in the Atlantic from 1975--2015: 
`View(dplyr::storms)`

Some other useful functions: 

* You can get a list of the names of
  a named vector, a named list, or the columns of a tibble or `data.frame`
  with the `names` function: `names(x)`, where x is a vector, list,
  tibble, or `data.frame`.

* You can get the length of a vector or list with the `length` function,
  and you can get the number of rows and columns in a tibble using the
  `dim` function:
  
```{r dim_len, include=TRUE}
x = c(1, 2, 3, 4, 5)
print("Length of x is")
print(length(x))

print("Dimensions of dplyr::storms is ")
dim(dplyr::storms)
```
That's `r nrow(dplyr::storms)` rows and `r ncol(dplyr::storms)` columns.
You can also get just the number of rows or the number of columns
with `nrow()` and `ncol()`.

# The Tidyverse 

The "tidyverse" is a collection of packages written by Hadley Wickham
to make it easy to work with data frames. Wickham developed an improved 
kind of data frame that has features that are lacking in the basic R
`data.frame`, and he developed a collection of tools for manipulating, 
analyzing, and graphing data from tibbles and regular `data.frames`.

To use the tidyverse, we need to load the package using R's `library` function.
If tidyverse is not installed on your computer, you will get an error message
and you will have to run `install.packages("tidyverse")` before you can 
proceed.

When you load `tidyverse`, it automatically loads a bunch of useful packages
for manipulating and analyzing data: `tibble`, `dplyr`, `tidyr`, `purrr`, 
`readr`, and `ggplot2`.

If you have the `pacman` package installed, it can help you avoid these error 
messages: after you load `pacman` with `library(pacman)`, then you can load
other packages using `p_load(tidyverse)` (you can substituate any other
package name for "tidyverse"): pacman will first see whether you have that
package on your computer; if you do, pacman will load it, and if you don't
pacman will install the package from the Comprehensive R Archive Network (CRAN)
and then load it.

In the code below, I will also load the `lubridate` package, which is part of
`tidyverse` but is not loaded automatically when you load `tidyverse`. 
`lubridate` provides useful functions for working with dates, which will
come in handy as we work with the weather data.

```{r load_tidyverse, include=TRUE}
library(tidyverse)
library(lubridate)
# alternately, I could do the following:
# library(pacman)
# p_load(tidyverse, lubridate)
```

One part of the `tidyverse` is the package `dplyr`, which has many useful 
tools for modifying and manipulating tibbles:

* `select` lets you choose a subset of columns from a tibble
* `rename` lets you rename columns
* `filter` lets you choose a subset of rows from a tibble
* `arrange` lets you sort the rows with respect to the values of different columns
* `mutate` lets you modify the values of columns or add new columns
* `summarize` lets you generate summaries of columns (e.g., the mean, maximum, 
  or minimum value of that column)
* `group_by` and `ungroup` let you perform calculations with grouping (e.g., in 
  combination with summarize, you can group by year to produce separate summaries 
  for each year)
* `bind_rows` to combine multiple tibbles that have the same kinds of columns
   by stacking one above the other.

There is a lot more, but these functions will be enough to keep us busy for now
and they will allow us to do some powerful analysis.

Let's start with `select`: You can select columns to keep or columns to delete.

Here are the first few rows of `nashville_weather`
```{r nw_head, include=TRUE}
head(nashville_weather)
```

Let's get rid of the `id` column, since we don't really care about the ID number,
that meteorological agencies use to identify the weather station. 
and let's get rid of the `location` column because we know that the data set
is from Nashville, so having that information repeated on each row is a waste
of space. 
To do this, we just  call select, specifying the tibble or `data.frame` to 
operate on, and then give a list of columns to eliminate, with a minus sign
in front of each:

```{r nw_no_id, include=TRUE}
x = select(nashville_weather, -id, -location)
head(x)
```

Alternately, instead of telling select which columns to get rid of, we can tell
it which columns to keep:

```{r nw_keep, include=TRUE}
x = select(nashville_weather, date, prcp, tmin, tmax)
head(x)
```

We can specify a range of consecutive columns by giving the first and last with a 
colon between them:

```{r nw_keep, include=TRUE}
x = select(nashville_weather, date:tmax)
head(x)
```

This is a general R thing: we can specify a range of numbers in a similar way:

```{r sequence, include=TRUE}
1:10
```

`rename` lets us rename columns:

```{r rename, include=TRUE}
x = rename(nashville_weather, weather_station = id, city = location)
head(x)
```

`filter` lets us select only rows that match a condition:
```{r filter_nw, include=TRUE}
x = filter(nashville_weather, year(date) > 2015 & tmax < 0)
head(x)
```
In the code above, I used the `year` function from the `lubridate` package to extract just the year from a date.

One thing that is important to know about making comparisons in `filter` expressions: to specify that two things are
equal, you write `==` with two equal signs. A single equal sign is for assigning a value to a variable and two
equal signs are for comparisons. You can also use "<=" for less than or equal to and ">=" for greter than or equal to.

```{r filter_nw_2, include=TRUE}
x = filter(nashville_weather, date == ymd("2016-06-10"))
x
```

In the code above, I used the `ymd` function from `lubridate` to translate a character
value "2016-06-10" to the date value for June 10, 2016.

You can combine conditions in `filter` by using `&` to indicate
"and" and `|` to indicate "or".


We can sort the rows of a tibble or `data.frame` with the `arrange` function:

```{r arrange, include=TRUE}
x = arrange(nashville_weather, desc(tmax), tmin)
head(x,10)
```

This sorts the rows in descending order of `tmax` (i.e., so the largest values are at the top),
and where multiple rows have the same value of `tmax`, then it sorts them in ascending order
of `tmin`. Observe the three rows with `tmax` = 41.7, the two rows where `tmax` = 41.1,
and the four rows where `tmax` = 40.6.

We can use `mutate` to modify the values of columns or to create new columns.
The temperatures in Nashville weather are in Celsius and the precipitation is
in millimeters. Let's convert these to Fahrenheit and inches, respectively, and
then let's create a `trange` column that will have the difference between the maximum
and minimum temperature:

```{r mutate, include=TRUE}
x = mutate(nashville_weather, prcp = prcp / 25.4, tmin = tmin * 9./5. + 32,
           tmax = tmax * 9./5. + 32, trange = tmax - tmin)
head(x)
```

Summaries are useful for finding averages and extreme values. Let's find the maximum and minimum temperatures 
and the most extreme rainfall in the whole data set:

```{r summarize_simple, include=TRUE}
x = summarize(nashville_weather, prcp.max = max(prcp), tmin.min = min(tmin), tmax.max = max(tmax))
x
```

`nashville_weather` has `r nrow(nashville_weather)` rows, but summarize reduces it to a single summary row.

You can use summary to generate multiple summary quantities from a column:

```{r summarize_simple_2, include=TRUE}
x = summarize(nashville_weather, prcp.max = max(prcp), prcp.min = min(prcp))
x
```

We can also generate grouped summaries:

```{r summarize_grouped_1, include=TRUE}
x = ungroup(summarize(group_by(nashville_weather, year(date)), prcp.max = max(prcp), prcp.tot = sum(prcp)))
head(x)
```
This provides the maximum one-day precipitation and the total annual precipitation for each year

Note how difficult it is to read that grouped summary expression: the `group_by` function is inside
`summarize`, which is inside `ungroup`. 

The `tidyverse` offers us a much nicer way to put these kind of complicated expressions together 
using what it calls the "pipe" operator, `%>%`.  The pipe operator chains operations together,
taking the result of the one on the left and inserting it into the one on the right.

We can use the pipe operator to rewrite the expression above as

```{r summarize_grouped_pipes, include=TRUE}
x = nashville_weather %>% group_by(year(date)) %>% summarize(prcp.max = max(prcp), prcp.tot = sum(prcp)) %>%
  ungroup()
head(x)
```
Now the expression is easier to read: First we group `nashville_weather` by year, then we summarize it by calculating the
maximum daily precipitation and the yearly total for each year, and finally, after we summarize we ungroup.

You can combine any set of the `tidyverse` functions using the pipe operator, so you could `select` columns, 
`filter` rows, `mutate` the values of columns, and `summarize`, using the `%>%` pipe operator to connect
all of the different operations in sequence.

The final `dplyr` command we're going to look at is `bind_rows`, which lets us combine tibbles:

```{r bind_rows, include=TRUE}
weather = bind_rows(nashville_weather, chicago_weather)
```
This creates a single tibble that has all of the rows from `nashville_weather` on 
top and all the rows from `chicago_weather` on the bottom. Because the two tibbles
have the same columns, the columns are matched up.

We can operate on this combined tibble:

```{r process_combined_data_frame}
x = weather %>% 
  mutate(year = year(date), t.range = tmax - tmin) %>%
  group_by(year, location) %>%   
  summarize(prcp.max = max(prcp), prcp.tot = sum(prcp), t.range.max = max(t.range)) %>%
  ungroup() %>%
  arrange(year, location)
head(x)
```
